Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-17 paper.qmd
Removed white space
figure captions in the two applications

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and (hyper-)~parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

- Remove all the extraneous commented text (Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible (changed in Rendering the model: Done)
- Check spelling is consistently American (Done)
- Check references (Done)
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place. (Done)
- Move the notation table to appendix (Done)
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.

@@

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Eight different NLDR representations of the same data. Different methods and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
#| fig-pos: H
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 30), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

The paper is organized as follows. @sec-background provides a summary of the literature on NLDR, and high-dimensional data visualization methods. @sec-method contains the details of the new methodology, including simulated data examples. In @sec-bestfit, we describe how to assess the best fit and identify the most accurate \gD{} layout based on the proposed model diagnostics. Curiosities and unexpected patterns discovered in NLDR results by examining the model in the data space are discussed in @sec-curiosities. Two applications illustrating the use of the new methodology for bioinformatics and image classification are in @sec-applications. Limitations and future directions are provided in @sec-discussion.

@@

<!-- - Connection between NLDR and MDS-->
Historically, low-dimensional (\kD{}) representations of high-dimensional (\pD{}) data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (A contemporary comprehensive guide to MDS can be found in @borg2005.)
<!-- Does JASA allow footnotes? If so this is a footnote, otherwise the punctuation has gone a bit rogue - can't have an entire sentence in brackets surely --> The \kD{} representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{} layout is constructed by minimizing a stress function that differences distances between points in \pD{} with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

<!-- Non-expert comment: This to me is a leap from MDS to NLDR. Is NLDR an MDS? I don't think so, but I'm missing the link -->
<!-- Suggested rewording: "To better capture specific structures potentially existing in \pD{}, many new methods for NLDR have emerged in recent years." -->
Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques: tSNE, UMAP, PHATE, TriMAP and PaCMAP. Both tNSE and UMAP can be considered to produce the \kD{} representation by minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@joreskog1969] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

@@

tsne_two_curvy_scaled <- read_rds("data/two_nonlinear/tsne_two_curvy_scaled.rds")
hex_grid_with_counts <- read_rds("data/two_nonlinear/hex_grid_with_counts.rds")
hex_grid_nonempty <- read_rds("data/two_nonlinear/hex_grid_nonempty.rds")
tr_from_to_df_two_curvy <- read_rds("data/two_nonlinear/tr_from_to_df_two_curvy.rds")
df_bin_centroids_two_curvy <- read_rds("data/two_nonlinear/df_bin_centroids_two_curvy.rds")

@@

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = emb1, y = emb2)) +
  geom_point(alpha = 0.5, colour = clr_choice, size = 0.5)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, colour = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 2)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             colour = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2) 
```

@@

trimesh_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             colour = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             colour = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

## Compute error
error_df_two_curvy_abs <- read_rds("data/two_nonlinear/error_df_two_curvy_abs.rds")

@@

@@

@@

## First projection
proj_obj1 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj1.rds")

@@

## Second projection

proj_obj2 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj2.rds")

two_curvy_proj_tsne_all_model1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5), 
  axis_text_size = 2, 
  is_category = FALSE) +
  interior_annotation(label = "c", position = sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)

@@

To illustrate the method, we use $7\text{-}D$ simulated data, which we call the "nonlinear clusters". It is constructed by simulating two clusters, each consisting of $1000$ observations. The C-shaped cluster is generated from $\theta \sim U(\text{-}3\pi/2, 0)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the C), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$, $X_4 = \cos(\theta)$. The other cluster is from $X_1 \sim U(0, 2)$, $X_2 \sim U(0, 3)$, $\gamma \sim U(0, 0.5)$, $X_3 = \text{-}(X_1^3 + X_2) + \gamma$, and $X_4 \sim U(0, 2)$. We would consider $T=(X_1, X_2, X_3, X_4)$ to be the geometric structure (true model) that we hope to capture.

@@

                 colour = "#bdbdbd") +

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and (hyper)parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

- Remove all the extraneous commented text (NOT Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible
- Check spelling is consistently American
- Check references
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place.
- Move the notation table to appendix
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.

@@

```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Eight different NLDR representations of the same data. Different techniques and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%
#| fig-pos: H
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 30), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

@@

The paper is organized as follows. @sec-background provides a summary of the literature on NLDR, and high-dimensional data visualization methods. @sec-method contains the details of the new methodology, including simulated data examples. Two applications illustrating the use of the new methodology for bioinformatics and image classification are in @sec-applications. Limitations and future directions are provided in @sec-discussion.

@@

<!-- - Connection between NLDR and MDS-->
Historically, \kD{} representations of \pD{} data have been computed using multidimensional scaling (MDS) [@kruskal1964], which includes principal components analysis (PCA) [@jolliffe2011] as a special case. (A contemporary comprehensive guide to MDS can be found in @borg2005.) The \kD{} representation can be considered to be a layout of points in \kD{}   produced by an embedding procedure that maps the data from \pD{}. In MDS, the \kD{} layout is constructed by minimizing a stress function that differences distances between points in \pD{} with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Challenges in working with high-dimensional data, including visualization, are outlined in @johnstone2009.

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD{}. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the \kD{}   minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)

```{r}
#| label: tbl-notation
#| tbl-cap: "Summary of notation for describing new methodology."
# Notation used in the paper

notation_df <- read_csv("misc/notation.csv")

# Create the table
kable(notation_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("3cm", "12cm"))
```

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

@@

```{r}
#| label: two-curvy-training
training_data_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_data.rds") |>
  mutate(ID = row_number())
```

```{r}
#| label: two-curvy-true-proj-data

true_model_df <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_true_model.rds")
wireframe_true_model <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_true_model_connections.rds")

data_two_curvy <- training_data_two_curvy |>
  select(-ID) |>
  mutate(type = "data")

true_model_two_curvy <- true_model_df |> 
  mutate(type = "true model")
```

<!--tSNE applied for Two curvy data-->
```{r}
#| label: tsne-two-curvy
tsne_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds") |>
  mutate(ID = row_number()) |>
  rename(c("emb1" = "tSNE1",
           "emb2" = "tSNE2"))
```

@@

num_bins_x_two_curvy <- 24

algo_obj_two_curvy <- fit_highd_model(
  highd_data = training_data_two_curvy, 
  nldr_data = tsne_two_curvy, 
  bin1 = num_bins_x_two_curvy, 
  q = 0.1, 
  benchmark_highdens = 1)

tsne_two_curvy_scaled <- algo_obj_two_curvy$nldr_obj$scaled_nldr
tr_from_to_df_two_curvy <- algo_obj_two_curvy$trimesh_data
df_bin_centroids_two_curvy <- algo_obj_two_curvy$model_2d
df_bin_two_curvy <- algo_obj_two_curvy$model_highd
hex_grid <- algo_obj_two_curvy$hb_obj$hex_poly
counts_df <- algo_obj_two_curvy$hb_obj$std_cts

hex_grid_with_counts <- left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hexID"))

hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids_two_curvy$hexID)

bin_width <- algo_obj_two_curvy$hb_obj$a1

@@

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = emb1, y = emb2)) +
  geom_point(alpha = 0.5, color = clr_choice, size = 0.5)

@@

# Second plot (identical to base except for xlim/ylim)
nldr_two_curvy2 <- nldr_two_curvy_original +
  interior_annotation("a", sc_ltr_pos, cex = 1.5)

# Third plot with aspect ratio and different annotation
nldr_two_curvy_original2 <- nldr_two_curvy_original +
  theme(aspect.ratio = 1) +
  interior_annotation("a1", sc_ltr_pos)

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, color = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 2)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims) 
```

@@

# df_bin_centroids_two_curvy <- df_bin_centroids_two_curvy |>
#   dplyr::filter(std_counts > 0)

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             color = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2) 
```

@@

trimesh_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             color = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)
```

@@

## Compute error
error_df_two_curvy_abs <- augment(
  highd_data = training_data_two_curvy,
  model_2d = df_bin_centroids_two_curvy,
  model_highd = df_bin_two_curvy)

error_df_two_curvy_abs <- error_df_two_curvy_abs |>
  mutate(sqrt_row_wise_total_error = sqrt(row_wise_total_error))

error_df_two_curvy_abs <- error_df_two_curvy_abs |>
  bind_cols(tsne_two_curvy_scaled |>
              select(-ID))

@@

# error_plot_two_curvy_hist <- 
#   ggplot(error_df_two_curvy_abs) +
#   geom_histogram(aes(x=sqrt_row_wise_total_error, y=..density..)) +
#   geom_density(aes(x=sqrt_row_wise_total_error, y=..density..), colour="red") +
#   #xlab(expression(group("|", e[hj], "|"))) +
#   xlab(expression(e[hj])) +
#   ylab("") +
#   theme_bw()

@@

```{r}
#| label: prep-data-for-proj-tsne-best

df_bin_two_curvy <- df_bin_two_curvy |>
  select(-hexID) |>
  mutate(type = "model")

# Apply the scaling
df_model_data_two_curvy <- bind_rows(data_two_curvy, true_model_two_curvy, df_bin_two_curvy)

scaled_two_curvy <- scale_data_manual(df_model_data_two_curvy, "type") |>
  as_tibble()

scaled_two_curvy_data <- scaled_two_curvy |>
  filter(type == "data") |>
  select(-type)

scaled_two_curvy_data_true_model <- scaled_two_curvy |>
  filter(type == "true model") |>
  select(-type)

scaled_two_curvy_data_model <- scaled_two_curvy |>
  filter(type == "model") |>
  select(-type)

```{r}
#| label: langevitour-true-two-curvy-proj
#| eval: false
df_model_data_two_curvy_filtered <- bind_rows(true_model_two_curvy, data_two_curvy)

langevitour::langevitour(df_model_data_two_curvy_filtered[1:(length(df_model_data_two_curvy_filtered)-1)],
                         lineFrom = wireframe_true_model$from,
                         lineTo = wireframe_true_model$to,
                         group = factor(df_model_data_two_curvy_filtered$type,
                                        c("data", "true model")),
                         levelColors = c(clr_choice, "#000000"))
```

```{r}
#| label: langevitour-best-fit-two-curvy-proj-tsne
#| eval: false

df_model_data_two_curvy_filtered <- bind_rows(df_bin_two_curvy, data_two_curvy)

langevitour::langevitour(df_model_data_two_curvy_filtered[1:(length(df_model_data_two_curvy_filtered)-1)],
                         lineFrom = tr_from_to_df_two_curvy$from,
                         lineTo = tr_from_to_df_two_curvy$to,
                         group = factor(df_model_data_two_curvy_filtered$type,
                                        c("data", "model")),
                         levelColors = c(clr_choice, "#000000"))
```

@@

## First projection
model_prj1 <- cbind(
  c(0.37847,-0.06517,0.04231,0.01655),
  c(-0.06466,-0.27052,0.22131,-0.15235))

proj_obj1 <- quollr::get_projection(projection = model_prj1, 
               proj_scale = 1, 
               highd_data = scaled_two_curvy_data, 
               model_highd = scaled_two_curvy_data_model, 
               trimesh_data = tr_from_to_df_two_curvy, 
               axis_param = list(limits = 0.5, 
                                 axis_scaled = 2,
                                 axis_pos_x = -0.5, 
                                 axis_pos_y = -0.5, 
                                 threshold = 0.03))

## For true model

projection_scaled <- model_prj1 * 1

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% projection_scaled

projected_true_model_df <- projected_true_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

true_model_df_proj <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_model_df, 
  by = c("from" = "ID"))

names(true_model_df_proj)[3:NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_from")

true_model_df_proj <- dplyr::left_join(true_model_df_proj, projected_true_model_df, by = c("to" = "ID"))
names(true_model_df_proj)[(2 + NCOL(projected_true_model_df)):NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_to")

proj_obj1_true <- proj_obj1
proj_obj1_true[["model_df"]] <- true_model_df_proj

two_curvy_proj_true <- plot_proj(
  proj_obj = proj_obj1_true, 
  point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.6, 0.6), 
  axis_text_size = 4, 
  is_category = FALSE) +
  interior_annotation(label = "a", cex = 1.5) +
  theme(aspect.ratio=1, 
        legend.position = "none")

# Changed the axis parameters
axis_obj <- gen_axes(
    proj = model_prj1 * 2,
    limits = 0.9,
    axis_pos_x = -0.4,
    axis_pos_y = -0.4,
    axis_labels = names(scaled_two_curvy_data),
    threshold = 0.05)

axes <- axis_obj$axes
circle <- axis_obj$circle

proj_obj1[["axes"]] <- axes
proj_obj1[["circle"]] <- circle

@@

## Second projection
model_prj2 <- cbind(
  c(-0.38281,-0.04870,0.00619,-0.02451),
  c(-0.01303,0.00634,-0.37421,0.09648))

proj_obj2 <- get_projection(projection = model_prj2, 
                            proj_scale = 1, 
                            highd_data = scaled_two_curvy_data, 
                            model_highd = scaled_two_curvy_data_model, 
                            trimesh_data = tr_from_to_df_two_curvy, 
                            axis_param = list(limits = 0.35, 
                                              axis_scaled = 2,
                                              axis_pos_x = -0.35, 
                                              axis_pos_y = -0.35, 
                                              threshold = 0.01))

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% model_prj2

projected_true_model_df <- projected_true_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

true_model_df_proj <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_model_df, 
  by = c("from" = "ID"))

names(true_model_df_proj)[3:NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_from")

true_model_df_proj <- dplyr::left_join(true_model_df_proj, projected_true_model_df, by = c("to" = "ID"))
names(true_model_df_proj)[(2 + NCOL(projected_true_model_df)):NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_to")

proj_obj2_true <- proj_obj2
proj_obj2_true[["model_df"]] <- true_model_df_proj

two_curvy_proj_true2 <- plot_proj(
    proj_obj = proj_obj2_true, 
    point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "black"), # linewidth, alpha
    plot_limits = c(-0.42, 0.42), 
    axis_text_size = 4, 
    is_category = FALSE) +
  interior_annotation(label = "b", cex = 1.5) +
  theme(aspect.ratio = 1, 
        legend.position = "none")

#Changed the axis parameters
axis_obj <- gen_axes(
    proj = model_prj2 * 2,
    limits = 0.7,
    axis_pos_x = -0.35,
    axis_pos_y = -0.35,
    axis_labels = names(scaled_two_curvy_data),
    threshold = 0.02)

axes <- axis_obj$axes
circle <- axis_obj$circle

proj_obj2[["axes"]] <- axes
proj_obj2[["circle"]] <- circle

two_curvy_proj_tsne_all_model1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5), 
  axis_text_size = 2, 
  is_category = FALSE) +
  interior_annotation(label = "c", position = sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1) #+
  #geom_segment(
  #  data = model_df, 
  #  aes(
  #    x = proj1_from, 
  #    y = proj2_from, 
  #    xend = proj1_to, 
  #    yend = proj2_to), 
  #  color = "#000000")

@@

To illustrate the method, we use $7\text{-}D$ simulated data, which we call the "non-linear clusters". It is constructed by simulating two clusters, each consisting of $1000$ observations. The C-shaped cluster is generated from $\theta \sim U(-3\pi/2, 0)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the C), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$, $X_4 = \cos(\theta)$. The other cluster is from $X_1 \sim U(0, 2)$, $X_2 \sim U(0, 3)$, $\gamma \sim U(0, 0.5)$, $X_3 = -(X_1^3 + X_2) + \gamma$, and $X_4 \sim U(0, 2)$. We would consider $T=(X_1, X_2, X_3, X_4)$ to be the geometric structure (true model) that we hope to capture.

@@

                 color = "#bdbdbd") +
